Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-03 doc/paper.Rmd
edits

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot,
    eda_plot,
    s_spei_plot,
    s_spei_diff_plot,
    g_spei_plot,
    g_spei_diff_plot,
    f_spei_plot,
    f_spei_diff_plot,
    size_plot,
    f_cf,
    f_1ha,
    model_data,
    xa_spei
  ))
})
```

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments---e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017]---can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

@@

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
For instance, <!--# parts of the Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50--100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable---particularly for plants, which are unable to move in response to changes in local conditions.
In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

<!--# Add here what is known about frag and drought effects - without setting it up so people expect a pval for an interaction term -->

@@

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900--2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

@@

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1--7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10 ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10 ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500--4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

@@

<!--# I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

<!--# re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the `mgcv` package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see @woodChapterSmoothers and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in @marraPracticalVariableSelection2011.

@@

```{r include=FALSE}
summary_stats <-
  model_data %>%
  group_by(habitat) %>%
  summarize(across(c(ht, shts, size, flwr, surv),
                   .fns = list(mean = mean, sd = sd),
                   na.rm = TRUE)) %>%
  right_join(model_data %>% count(habitat)) %>%
  mutate(across(ends_with("_mean") | ends_with("_sd"),
                                ~round(.x, 2)))
summary_stats <- 
  full_join(
  summary_stats,
  model_data %>%
    count(habitat, plot) %>%
    group_by(habitat) %>%
    summarize(n_mean = mean(n), n_sd = sd(n)) %>%
    mutate(across(where(is.numeric), round))
)
summary_stats

by_year <-
  model_data %>% 
  group_by(year, habitat) %>% 
  summarize(prop_surv = mean(surv)) %>% 
    arrange(prop_surv)
by_year

xa_spei %>% ungroup() %>%
  summarize(across(everything(), mean)) %>%
  filter(year(yearmonth)==2005)

Our dataset comprised `r summary_stats$n[2]` plants in continuous forest and `r summary_stats$n[1]` plants in forest fragments, with plots in CF having on average 2.6-fold more plants than plots in 1-ha fragments (CF = `r summary_stats$n_mean[2]` ± `r summary_stats$n_sd[2]` 1-ha = `r summary_stats$n_mean[1]` ± `r summary_stats$n_sd[1]` ).
When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (P(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in December 2005 and January 2006.
Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average \~`r (1 - (2.5/2.9))*100 %>% round()`% fewer shoots (2.5 ± 1.5 SD) and were \~`r (1 - (36.2/40.6))*100 %>% round()`% shorter (36.3 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently substantially larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (`r summary_stats$flwr_mean[2]` ± `r summary_stats$flwr_sd[2]` vs. `r summary_stats$flwr_mean[1]` ± `r summary_stats$flwr_sd[1]`, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold size of about 148 [log(size) \> 5] (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots <!--# not sure I can/should fill in that blank -->.
Even though flowering probability of large plants [log(size) \> 5] was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

@@

@@

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 13.1 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--11 (March--June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.

@@

FInally, with the exception of survival in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
<!--# more here -->.
Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
Using a decadal dataset of plant demography, we show that some---but not all---results from short-term studies can be extrapolated to longer times scales.
Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).
These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance---populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, these differences in flowering probability could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets.
    Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon.
    Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought.
    Dry season droughts result in green-up, but not wet season droughts.
    During drought, old leaves are shed and new leaves are maintained.
    2-5 months for newly flushed leaves to fully expand.

>>>>

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>",
	dpi = 300
)
library(targets)
library(knitr)
library(here)
```

```{r load-targets}
withr::with_dir(here(), {
  tar_load(c(
    normals_plot,
    eda_plot,
    s_spei_plot,
    s_spei_diff_plot,
    g_spei_plot,
    g_spei_diff_plot,
    f_spei_plot,
    f_spei_diff_plot,
    size_plot,
    f_cf,
    f_1ha
  ))
})
```

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015].
Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat.
Although the local extinction of species from individual fragments is common [@haddadHabitatFragmentationIts2015]<!--# others? -->, the demographic mechanisms underlying these extinctions are poorly understood [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments -- e.g., increased air temperature, decreased relative humidity, reduced soil moisture, increased solar radiation [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999; @arroyo-rodriguezDoesForestFragmentation2017] -- can depress rates of reproduction or survivorship [@zartmanDroughtImpactsSurvivorship2015; @brunaSeedGerminationRainforest1999; @lauranceRainForestFragmentation1998], leading to local or regional population declines [@brunaHabitatFragmentationPlant2009]<!--# others? -->.
This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates.
For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in their extensive fragmentation and the creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].
Despite this, evidence linking changes population-level demographic responses with changes in environmental conditions remains scarce.

@@

Populations in fragmented tropical landscapes are also at risk from climate change [@brodieClimateChangeTropical2012; @scheffersExtremeThermalHeterogeneity2017], which can have complex or even contrasting effects across the same tropical biome.
For instance, <!--#parts of th      e Amazon basin have experienced a drying trend since the mid-1970's, and --> climate models for the next 50-100 years predict that some parts of Amazonia will experience decreases in dry season precipitation [@malhiClimateChangeDeforestation2008] and increases in the frequency and severity of droughts [@caiIncreasingFrequencyExtreme2014; @zengCausesImpacts20052008; @moraProjectedTimingClimate2013].
These changes could exacerbate the altered environmental conditions in fragments thought to make populations demographically vulnerable - particularly for plants, which are unable to move in response to changes in local conditions.
In contrast, other parts of the basin are predicted to see an increase in the frequency and severity of extreme precipitation events, which can also have detrimental consequences for plant survival and reproduction [@estebanOtherSideDroughts2021].

Add here what is known about frag and drought effects - without setting it up so people expect a pval for an intearaciton tern

@@

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

@@

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004].
Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = 1.1 flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (**Supplemental\_\_** <!--# insert fig ref -->).
This study uses data from 1-ha fragments and Continuous Forest sites.

@@

<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the *dlnm* package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recovered from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

<!-- re 36 mos transpalnt shock justification; how are lag lengths typically chosen?  -->

The crossbasis function was fit to the data in the context of a generalized additive model (GAM) with restricted maximum likelihood using the *mgcv* package [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see Wood [-@woodChapterSmoothers] and Gasparrini et al. [-@gasparriniPenalizedFrameworkDistributed2017] for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of months) and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal when most complex <!--# You may want to say what the 3 knots can capture: increasing then decreasing, exp decline etc -->.
Because of penalization, the number of knots is generally not important as long it is great enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution and the natural log of height $\times$ shoot number as our proxy for plant size.
A scaled t family was chosen because residuals were leptokurtic with a Gaussian error structure.
For reproduction in year t our response variable was whether or not a plant had any inflorescences (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function .

In the process of fitting the models, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity [@woodGeneralizedAdditiveModels2017].
We applied an additional penalty to shrink linear portions toward zero with the `select=TRUE` option to the `gam()` function, and inferred statistical significance of model terms with p-values from the `summary.gam()` function as recommended in Marra and Wood [-@marraPracticalVariableSelection2011].

@@

**Our dataset comprised XXXX plants in continuous forest and XXXX plants in forest fragments, with an plots in CF having on average X-fold more plants than plots in 1-ha fragments (mean per plot + SD in each) .** When summarizing across years and plots, the proportion of *Heliconia acuminata* that survived in CF and 1-ha was similarly high (P(surv) = 0.95).
The proportion of surviving plants was lowest in the 2003-2004 transition year (P(surv) = 0.92), which coincided with droughts in both the 2003 and 2004 rainy seasons (Figure \@ref(fig:eda)b,d).
The lowest survival for 1-ha fragment plots (p(surv) = 0.93) was for the 2005-2006 transition year, which encompassed a moderate drought in October 2005 and and wetter than average conditions (SPEI \> 0.5) in DOUBLE CHJECK DATES OF DROUGHT AND RAIN. Survival was size dependent in both continuous forest and 1-ha fragments (p \< 0.001 for the log-transformed effect of plant size in year t on suvival in year t+1).
While the survival probability of large plants approached 1 in both habitat types (Figure \@ref(fig:size-plot)a), the survival of the smallest plants was higher in 1-ha fragments.
However, the 95% confidence intervals for 1-ha fragments and continuous forest overlapped for all sizes.

<!--# The proportion of surviving plants was lowest from 2004-2005 (0.92), which coincided with a moderate drought in Jan 2004 / Feb 2005 and a severe drought in Jan–Mar of 2004 (these dates overlap? -->

<!--# This is across the whole timeseries, which I'm not sure is actually that important. What summary statistics would be useful? -->

Plants in continuous forest had an average of 2.9 shoots (± 1.8 SD) and were on average 40.6 cm tall (± 26.5 SD).
Plants in 1-ha fragments had on average **\~XX% fewer** shoots (2.5 ± 1.5 SD) and **were XX% shorter** (36.2 cm ± 24.1 SD).
Because the proxy for plant size used in our models was the product of these two metrics, plants in fragments were consequently significantly larger than those in forest fragments (150 ± 175 SD vs. 112 ± 141 SD, respectively).
This disparity in plant size was most pronounced in the initial years of our surveys and then diminished over time (Figure \@ref(fig:eda)a), likely due to a combination of greater seedling recruitment in CF and plant mortality or shrinkage As with survival, size in year t was a significant predictor of size in year t+1 (p \< 0.001).
While the effect was generally similar across size classes and habitat types, the impact of plant size on growth was greatest for mid-sized plants in continuous forest (Figure \@ref(fig:size-plot)b).

@@

While the overall proportion of plants flowering was very low, it was nevertheless almost 40% higher in continuous forest than 1-ha fragments (0.050 ± 0.021 vs. 0.036 ± 0.027, respectively).
This was true in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
This disparity was largely due to the fact that flowering is size-dependent (p \< 0.001), with the probability of flowering increasing dramatically once plants reached the threshold value of log 5\_\_\_\_ (Figure \@ref(fig:size-plot)c), which corresponds roughly to \_\_\_\_ shoots.
Even though flowering probability of large plants (log(size) \> 5) was greater in 1-ha fragments compared to continuous forest, there were proportionally far fewer plants above the reproductive threshold size in forest fragments (Figure \@ref(fig:size-plot)d).
Indeed, the most striking difference between habitat types coincided with a severe drought in 2003, when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest vs. only 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is. EB: same-->

@@

<!--# There was a pronounced threshold in flowering probability corresponding with a log size of about 5...is this 5 shoots? What abvout the proxy size? -->

@@

Drought history also had a significant effect on plant size in both habitats (p \< 0.001).
The fitted crossbasis functions for fragments and continuous forest showed somewhat similar patterns, although the surface for 1-ha fragments was much more complex (edf = 17.9 for 1-ha fragments; edf = 5.9 for continuous forest; see also Figure \@ref(fig:growth-spei)).
Drought at lags 8--11 (March - June, the end of the wet season) increase growth in both habitats, but more strongly in 1-ha fragments.

@@

FInally, with the exception of growth in 1-ha fragments (p = 0.257), the delayed effects of SPEI on all three vital rates varied significantly between plots (p \< 0.001 for the random effect of plot).

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
<!--# more here -->.
Despite this, surprisingly few studies have compared the structure and demography of populations in habitat fragments and continuous habitat; those that have are typically short term (i.e., \<4 years).
Using a decadal dataset of plant demography, we show that some - but not all - results from short-term studies can be extrapolated to longer times scales.
Our results also add to the emerging body of work showing delayed effects of climate drivers on demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
Finally, we show for the first time that the magnitude, direction, and temporal lag in demographic responses to drought differs between forest fragments and continuous forest.

@@

Decadal comparisons of *Heliconia acuminata* growth and survival largely mirrored those of prior experiments and surveys [@brunaHabitatFragmentationDemographic2002].
In contrast, conclusions regarding the effect of fragmentation on flowering differed substantially from those based on short-term surveys [@brunaHabitatFragmentationDemographic2002].
While in some years the probability a plant flowered was far lower in fragments than in continuous forest, in others it was comparable or even higher.
The probability that individual plants flowered continued to be size-dependent, with a similar size threshold below which plants were unlikely to produce inflorescences.
However, while in our initial three-year survey the relationship between plant size and probability of flowering was similar in both habitat types [@brunaHabitatFragmentationDemographic2002], when using the decadal dataset were found that flowering probability for the largest plants is actually higher in forest fragments (i.e., average fitted probability of flowering for plants in upper size quartile = 0.125 in CF vs. 0.094 in FF).

These size-dependent differences in flowering probability had only a modest effect on the previously observed disparities in flowering plant abundance - populations in fragments are smaller, and they have proportionately fewer of these large plants [@brunaHabitatFragmentationDemographic2002].
However, they could potentially lead to major habitat-specific differences in lifetime reproductive effort for these long-lived plants, as they translate into flowering once every 7.97 years in fragments vs. every 10.63 years in continuous forest.
Our results underscore why caution is needed when using short-term studies to draw conclusions regarding the effects of fragmentation on long-lived organisms: seemingly small (i.e., non-significant) differences in vital rates, when compounded over multiple years, can lead to demographic divergence among populations [@gagnonGrowthUnderstoryHerb2011].

@@

-   [@janssenDroughtEffectsLeaf2021] Used machine learning to combine remote sensing and ground surveys of literfall from published datasets. Estimates green-up, literfall, growth, etc. for central and southern america including the Amazon. Drought results in anomolously high leaf flushing at start of drought followed by increased litterfall durring the drought. Dry season droughts result in green-up, but not wet season droughts. During drought, old leaves are shed and new leaves are maintained. 2-5 months for newly flushed leaves to fully expand

Effects beyond Ha
